Solving Graph Homomorphism and Subgraph Isomorphism Problems Faster Through Clique Neighbourhood Constraints

### 2021-05-27 paper/paper.tex
Editorial tweaks

\paragraph{Degree-based invariants.} Constraint programming techniques for finding
homomorphisms can be made much more effective by exploiting certain invariants. It is
straightforward to verify the following.

\begin{proposition}[neighbourhood degree sequences are preserved]\label{proposition:degreends}
    Suppose  $i$ is a locally injective homomorphism from $G$ to  $H$. Let
    $v$ be any vertex. Then $i$ cannot map $v$ to a vertex of lower degree, $\degree_G(v) \le \degree_H(i(v))$.
    Furthermore, $v$ cannot be mapped to any vertex whose neighbourhood degree sequence is
            not pointwise greater than or equal to its own.
\end{proposition}
Since subgraph isomorphisms are in particular locally injective, these invariants also hold for
subgraph isomorphisms \cite{DBLP:journals/constraints/ZampelliDS10}; indeed, every recent constraint
programming approach for subgraph isomorphism finding makes use of degrees and neighbourhood degree
sequences. The simplest way for a constraint programming algorithm to use these results is as unary
constraints, which are propagated when domains are initialised, before search begins. For each
domain representing a domain vertex, any value representing a codomain vertex whose degree or
neighbourhood degree sequence is too low may immediately be rejected.

@@

However, these invariants do \emph{not} hold for homomorphisms in general. As a
counter-example, a homomorphism may map a star graph onto a single edge, as in
\cref{figure:counterexamples} (top left).

\paragraph{Path-based invariants.} Constraint programming solvers can also exploit
invariants that are based upon paths. This is done by automatically adding \emph{implied}
constraints to the problem that are implied by the original model, but that will give stronger
propagation power.

@@

\paragraph{Cliques and homomorphisms.} We now discuss the main new technique of this paper. Observe
that so far, we have not seen any unary constraints which are valid for the general homomorphism
problem. All existing constraint programming approaches for the subgraph isomorphism problem begin
by branching on the variable which has the smallest domain---typically this will be a domain for a
vertex of high degree, or which has many high degree neighbours. Furthermore, once one such
guessed assignment has been made, adjacency propagation means many other domains will be
substantially reduced in size, making subsequent branching choices simpler.  However, for
homomorphism problems, if we cannot find any implied unary constraints then every domain will
initially be of the same size, which will make it much harder for a solver to know where to start.
This motivates the following.

@@

Given the complexity of modern solvers, a critical question is how we can be sure that they are
producing correct answers. The Glasgow Subgraph Solver's subgraph isomorphism and clique algorithms
are both \emph{certifying} \cite{DBLP:journals/csr/McConnellMNS11}: that is, they can output a
mathematical proof that they have reached a correct answer by sound reasoning, and this proof log
can be verified by the VeriPB proof checking tool
\cite{DBLP:conf/cp/GochtMMNPT20,DBLP:conf/ijcai/GochtMN20}. We therefore extended the solver's
existing proof logging support to cover our additions. For the changes not involving clique-finding,
this was entirely routine---although proof logging did catch an implementation bug where we were
incorrectly applying degree-based filtering on the $G^{2,n}$ graphs for locally injective
homomorphisms, in violation of the counter-example to \cref{corollary:lishapes}. For clique-based
filtering, the process involved substantially more technical work. Briefly, for a given domain
vertex $v$ that cannot be mapped to a codomain vertex $w$, it is possible to instruct the proof
verifier to derive a set of conditional clique constraints saying that ``either $v$ cannot be mapped
to $w$, or the following clique-like problem must be satisfiable'' by following steps similar to the
maximum common subgraph to clique reduction presented by \citet{DBLP:conf/cp/GochtMMNPT20}, and then
to reuse clique proof-logging techniques to show that this clique-like problem is unsatisfiable.  We
were thus able to produce and verifying proofs of correctness for medium-sized instances, giving us
confidence our implementation was correct.

>>>>

\subsection{Invariants}

Constraint programming techniques for finding homomorphisms can be made much more effective by
exploiting certain invariants. It is straightforward to verify the following.

\begin{proposition}[neighbourhood degree sequences are preserved]\label{proposition:degreends}
    Suppose  $i$ is a locally injective homomorphism from $G$ to  $H$. Let
    $v$ be any vertex. Then $i$ cannot map $v$ to a vertex of lower degree, $\degree_G(v) \le \degree_H(i(v))$.
    Furthermore, $v$ cannot be mapped to any vertex whose neighbourhood degree sequence is
            not pointwise greater than or equal to its own.
\end{proposition}
Since subgraph isomorphisms are in particular locally injective, these invariants also hold for
subgraph isomorphisms \cite{DBLP:journals/constraints/ZampelliDS10}; indeed, every recent constraint
programming approach for subgraph isomorphism finding makes use of degrees and neighbourhood degree
sequences. The simplest way for a constraint programming algorithm to use these results is as unary
constraints, which are propagated when domains are initialised, before search begins. For each
domain representing a domain vertex, any value representing a codomain vertex whose degree or
neighbourhood degree sequence is too low may immediately be rejected.
However, these invariants do \emph{not} hold for homomorphisms in general. As a
counter-example, a homomorphism may map a star graph onto a single edge, as in
\cref{figure:counterexamples} (top left).

@@

Constraint programming solvers can also exploit invariants that are based upon paths. This is done
by automatically adding \emph{implied} constraints to the problem that are implied by the original
model, but that will give stronger propagation.

@@

We now discuss the main new technique of this paper. Observe
that so far, we have not seen any unary constraints which are valid for the general homomorphism
problem. All existing constraint programming approaches for the subgraph isomorphism problem begin
by branching on the variable which has the smallest domain---typically this will be a domain for a
vertex of high degree, or which has many high degree neighbours. Furthermore, once one such
guessed assignment has been made, adjacency propagation means many other domains will be
substantially reduced in size, making subsequent branching choices simpler.  However, for
homomorphism problems, if we cannot find any implied unary constraints then every domain will
initially be of the same size, which will make it much harder for a solver to know where to start.
This motivates the following.

@@

Given the complexity of modern solvers, a critical question is how we can be sure that they are
producing correct answers. The Glasgow Subgraph Solver's subgraph isomorphism and clique algorithms
are both \emph{certifying} \cite{DBLP:journals/csr/McConnellMNS11}: that is, they can output a
mathematical proof that they have reached a correct answer by sound reasoning, and this proof log
can be verified by the VeriPB proof checking tool
\cite{DBLP:conf/cp/GochtMMNPT20,DBLP:conf/ijcai/GochtMN20}. We therefore extended the solver's
existing proof logging support to cover our additions. For the changes not involving clique-finding,
this was entirely routine---although proof logging did catch an implementation bug where we were
incorrectly applying degree-based filtering on the $G^{2,n}$ graphs for locally injective
homomorphisms, in violation of the counter-example to \cref{corollary:lishapes}. For clique-based
filtering, the process involved substantially more technical work. Briefly, for a given domain
vertex $v$ that cannot be mapped to a codomain vertex $w$, it is possible to instruct the proof
verifier to derive a set of conditional clique constraints saying that ``either $v$ cannot be mapped
to $w$, or the following clique-like problem must be satisfiable'' by following steps similar to the
maximum common subgraph to clique reduction presented by \citet{DBLP:conf/cp/GochtMMNPT20}, and then
to reuse clique proof-logging techniques to show that this clique-like problem is unsatisfiable.  We
were able to produce and verify proofs of correctness for medium-sized instances, giving us
confidence in our implementation.
